Understanding brain complexity in naturalistic processing in humans

### 2023-02-13 paper/main.tex

We're interested in the complexity of brain patterns that underly
different types of thoughts. To explore this question space, we take
brain patterns recorded under different experimental conditions and
project them into lower dimensional spaces using principle components
analysis. We can then ask how well those low-dimensional embeddings of
the data retain cognitively relevant information like when in a story
someone is listening to. 
Then we can explore decoding accuracy as a
function of the number of components, or dimensions, in the
low-dimensional space under different cognitive conditions.

@@

This idea can be explored in this visual analogy (Fig.~\ref{fig: pie_example}) for neural
compression.  Here there are two images of pies, the top pie is more
complex than the bottom.  On the left we're illustrating that it takes
fewer componenets to reach the same 95 percent variance explained in
the less complex pie, which corresponds to higher compression.
However, on the right with very few components similar
variance is explaining both pies.

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{figs/pie_example.pdf}
  \caption{\textbf{Illustration of compression.} Visual analogy for
    neural compression. Here are 2 images of pies, one more complex
    thantheother. \textbf{a.}It takes fewer components to reach the same percent
    variance explained in the less complex pie, which corresponds to
    higher compression. \textbf{b.}However, with very few components, similar variance is explained in both pies.
\textbf{c.} Plots the cumulative explained variance for more and more components.}
  \label{fig:pie_example}
\end{figure}

@@

\section*{Methods}

\subsubsection*{Dimensionality reduction}

Explain PCA

\subsection*{Evaluation metrics}
 We evaluated the degree of
compression of held-out
neuroimaging data with the time at which it was collected.  We used this latter evaluations (using timepoint
decoding) as a proxy for
gauging how much explanatory power the compressed data held
with respect to the observed data.

@@

To explore how higher-order structure varies with stimulus structure
and complexity, we used a previous neuroimaging dataset
~\cite{SimoEtal16}  in which participants listened to an audio recording of a story; 36 particpants listen to an intact version of the story, 17 participants listen to time-scrambled recordings of the same story where paragraphs were scrambled, 36 particpants listen to word-scrambled version and 36 participants lay in rest condition.

Prior work has shown participants share similar neural responses to
richly structured stimuli when compared to stimuli with less
structure.  To assess whether the moment-by-moment higher order
correlations were reliably preserved across participants, we used
inter-subject functional connectivity (ISFC) to isolate the
time-varying correlational structure (functional connectivity patterns
that were specifically driven by the story participants listened to.
Following the analyses conducted by (HTFA)~\cite{MannEtal18}, we first
applied \textit{hierarchical topographic factor analysis} (HTFA) to
the fMRI datasets to obtain a time series of 700 node activities for
every participant.  We then applied dimensionality reduction
(Incremental PCA) for each group.

We then compared the groups’ activity patterns (using Pearson
correlations) to estimate the story times each corresponding pattern
using more and more principle components.

To assess decoding accuracy, we randomly divided participants for each
stimulus into training and testing groups. We then compared the
groups’ activity patterns (using Pearson correlations) to estimate the
story times each corresponding pattern using more and more principle
components (as the data became less compressed). Specifically, we asked, for each timepoint: what are the correlations
between the first group's and second group's activity patterns at each
order. We note that the decoding test we used is a conservative in which we count a timepoint label as incorrect if it is not an exact match.

\section*{Results}

For our decoding analysis, we used HTFA-derived node activities
~\cite{MannEtal18} from fMRI data collected as participants listened
to an audio recording of a story (intact condition; 36 participants),
listened to time scrambled recordings of the same story (17
participants in the paragraph-scrambled condition listened to the
paragraphs in a randomized order and 36 in the word-scrambled
condition listened to the words in a randomized order), or lay resting
with their eyes open in the scanner (rest condition; 36
participants). We sought to understand the 'dimensionality' of the
neural patterns by using more and more principle
components to decode a multi-subject fMRI datasets. This story listening dataset was
collected as part of a separate study, where the full imaging
parameters, image preprocessing methods, and experimental details may
be found~\citep{SimoEtal16}. The dataset is available at
\href{url}{http://arks. princeton.edu/ark:/88435/dsp015d86p269k}.

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{figs/decode_interpret.pdf}
  \caption{\textbf{Decoding accuracy.} \textbf{a. Decoding accuracy by
      number of components.} Ribbons of each color display
    cross-validated decoding performance for each condition (intact,
    paragraph, word, and rest). Decoders were trained using
    increasingly more principle components and displayed relative to
    chance (red line). \textbf{b. Fixed decoding accuracy by number of
      components.} We zoom in on the plot shown in \textbf{a.} and add
    a line denoting fixed decoding accuracy (.05). We plot where the
    intact, paragraph, and word conditions intersect.
    \textbf{c. Explanation of inflection metric.} First the we fit a sigmoid function to the decoding accuracy by number of components. Second, we found where the second derivate is both positive and less than .0001. Last, we then plot that inflection point as a single metric to capture the slope and asymptote of the curve.}
  \label{fig:decode_interpret}
\end{figure}

We performed a decoding analysis, using cross validation to estimate
(using other participants’ data) which parts of the story the additional added
principle component corresponded to (see \textit{Materials and methods}). We note that our primary goal was not to achieve perfect decoding accuracy, but rather to use decoding accuracy as a benchmark for assessing whether different neural features specifically capture cognitively relevant brain patterns.

Separately for each experimental condition, we divided participants
into two groups. Starting with 1 principle comonent, for
each dimension we added another principle component, and we correlated the group 1 activity patterns with group 2
activity patterns.  There were 272
timepoints for paragraph condition, 300 timepoints for intact and word
conditions, and 400 timepoints for rest condition,  so chance
performance on this decoding test is was $\frac{1}{272}$,
$\frac{1}{300}$, and $\frac{1}{400}$ respectively.

- As complexity of the stimulus increases, decoding accuracy
increases (Fig.~\ref{fig:decode_interpret},  a.).  Replication of
Simony findings.

- As complexity of the stimuli increases, we need fewer components to
decode the same amount (Fig.~\ref{fig:decode_interpret},  b.).

- As complexity of the stimuli increases, more components are required to reach peak decoding accuracy (Fig.~\ref{fig:decode_interpret},  c.).

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{figs/decode_pcs_network.pdf}
  \caption{\textbf{Inflection points by network.} \textbf{a.}
    Inflection point was calculated as explained in
    Fig.~\ref{fig:decode_interpret}, b. Analyses were limited by the
    brain networks (using the Yeoetal11 network parcellation) and
    arranged in increasing order relative to the intact condition.
    \textbf{b. and c.} For the total time in the intact condition, we are plotting the relative inflection points (\textbf{b.}) and corresponding number of components (\textbf{c.}) by network. \textbf{d.} The network parcellation defined by Yeo et al. (2011) is displayed on the inflated brain maps. The colors and network labels serve as a legend for \textbf{a.} and \textbf{d.}}
  \label{fig:decode_pcs_network}
\end{figure}

As complexity of the stimuli increases, decoding accuracy increases
with higher cognitive areas. (Fig.~\ref{fig: decode_pcs_network}).

 If, there is some understanding of the narrative that accumulates over time, we should be able to see that change.

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{figs/decode_pcs.pdf}
  \caption{\textbf{Inflection points by thirds.} \textbf{a.}
    Decoding accuracy by number of components not broken into thirds
    (Fig.~\ref{fig:decode_interpret} a.). \textbf{b. and c.}
    Quantifying changes in decoding accuracy across
time. \textbf{b.} Slope of decoding accuracy was calculated by fitting
a regression line for each component/condition for each
third. \textbf{c.}  We also repeated the analysis
(Fig.~\ref{fig:decode_interpret}, b.) to obtain the inflection point
for each condition and for each third. \textbf{d.}  Decoding accuracy by number of components for each third of the scan time. We repeated the same analysis in Fig.~\ref{fig:decode_interpret} a. but breaking the scan time for each condition into 3 intervals.
}
  \label{fig: decode_pcs_thirds}
\end{figure}

- Increases in decoding accuracy with the same number or fewer components for more complex, cognitively rich, conditions.
- Decreases in decoding accuracy for the word-scrambled and rest
condition.

@@

\section*{Author contributions}
Concept: J.R.M.  Implementation: T.H.C., L.L.W.O., and
J.R.M.  Analyses: L.L.W.O and J.R.M.

\bibliographystyle{apacite}
\bibliography{CDL-bibliography/memlab}

>>>>

We're interested in the complexity of brain patterns that underly
different types of thoughts. To explore this question space, we will take
brain patterns recorded under different experimental conditions used in Aim 2, and
project them into lower dimensional spaces using principle components
analysis. We can then ask how well those low-dimensional embeddings of
the data retain cognitively relevant information like when in a story
someone is listening to.

This work has been inspired, in part, by~\cite{MackEtal20}. In this paper, 
they investigated the role of the prefrontal cortex in filtering out 
irrelevant content.  Specifically, they looked at if the vmPFC performs 
data reduction on incoming information through compression.  
This was motivated, in part, by orbital frontal cortex (OFC) 
compression in rats~\citep{ZhouEtal19}. They studied this using a learning paradigm in which participants had to classify insects based on different numbers of feature dimensions.  The idea was that participants in some learning blocks, participants could identify the insects based on one feature (low complexity) or several features (high complexity), but importantly the stimuli remained the same across all learning problems.  They found that complexity and compression had an inverse relationship; the lower complexity of a conceptual space, the higher the degree of compression.  Building on this idea, we wonder if varying degrees of compression is performed throughout the brain.  We also want to test this idea, but using varying levels of engagement listening to a naturalistic stimuli.

To understand the degree of compression throughout the brain during cognition, we will use the same fMRI data from Aim 2, collected while participants listened to a story in different scrambling conditions.  We will measure the degree that multivoxel activation patterns are compressed during story listening using principle components analysis (PCA) a method for low-rank approximation of multidimensional data~\citep{EckaYoun36}. We will explore this using decoding accuracy as a function of the number of components, or dimensions, in the low-dimensional space under different cognitive conditions.

@@

This idea can be explored in this visual analogy (Fig.~\ref{fig: pie_example}) for neural compression.  Here there are two images of pies, the top pie is more complex than the bottom.  On the left we're illustrating that it takes
fewer components to reach the same 95 percent variance explained in
the less complex pie, which corresponds to higher compression.
However, on the right with very few components similar
variance is explaining both pies.

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{figs/pie_example.pdf}
  \caption{\textbf{Illustration of compression.} Visual analogy for
    neural compression. Here are 2 images of pies, one more complex
    than the other. \textbf{a.} It takes fewer components to reach the same percent
    variance explained in the less complex pie, which corresponds to
    higher compression. \textbf{b.} However, with very few components, similar variance is explained in both pies.
\textbf{c.} Plots the cumulative explained variance for more and more components.}
  \label{fig:pie_example}
\end{figure}

@@

% PCA, or the closely related factor analysis and singular value decomposition (SVD) (Hastie et al., 2009), is widely used in the study of individual differences and aids estimating how many latent components, or “factors”, underlie a pattern of (item) responses within or across participants, as for instance in the context of intelligence (Spearman, 1904) or personality tests (Cattell, 1947). In the context of neuroimaging, PCA has been used to identify brain networks (Huth et al., 2012; Friston et al., 1993).

 % add citation:
% S.M.Smith,A.Hyva ̈rinen,G.Varoquaux,K.L.Miller,and C. F. Beckmann, “Group-PCA for very large fMRI datasets,” NeuroImage, vol. 101, pp. 738, 2014.

\subsection*{Evaluation metrics}
 We will evaluate the degree of
compression of held-out
neuroimaging data by assessing the time at which it was collected.  We will use this evaluation (timepoint
decoding) as a proxy for
gauging how much explanatory power the compressed data held
with respect to the observed data.

@@

To explore how compression varies with complexity, we will use a previous neuroimaging dataset
~\cite{SimoEtal16}  in which participants listened to an audio recording of a story; 36 particpants listen to an intact version of the story, 17 participants listen to time-scrambled recordings of the same story where paragraphs were scrambled, 36 particpants listen to word-scrambled version and 36 participants lay in rest condition.

Following the analyses conducted by (HTFA)~\cite{MannEtal18}, we first
apply \textit{hierarchical topographic factor analysis} (HTFA) to
the fMRI datasets to obtain a time series of 700 node activities for
every participant.  We then apply dimensionality reduction
(Incremental PCA) for each group.

We then compare the groups’ activity patterns (using Pearson
correlations) to estimate the story times each corresponding pattern
using more and more principle components.

To assess decoding accuracy, we randomly divide participants for each
stimulus into training and testing groups. We then compare the
groups’ activity patterns (using Pearson correlations) to estimate the
story times each corresponding pattern using more and more principle
components (as the data became less compressed). Specifically, we ask, for each timepoint: what are the correlations
between the first group's and second group's activity patterns at each
order. We note that the decoding test we used is a conservative in which we count a timepoint label as incorrect if it is not an exact match.

% \subsection*{Evaluation metrics}
%  We evaluated the degree of
% compression of held-out
% neuroimaging data with the time at which it was collected.  We used this latter evaluations (using timepoint
% decoding) as a proxy for
% gauging how much explanatory power the compressed data held
% with respect to the observed data.

% \subsubsection*{Timepoint decoding}

% To explore how higher-order structure varies with stimulus structure
% and complexity, we used a previous neuroimaging dataset
% ~\cite{SimoEtal16}  in which participants listened to an audio recording of a story; 36 particpants listen to an intact version of the story, 17 participants listen to time-scrambled recordings of the same story where paragraphs were scrambled, 36 particpants listen to word-scrambled version and 36 participants lay in rest condition.

% Prior work has shown participants share similar neural responses to
% richly structured stimuli when compared to stimuli with less
% structure.  To assess whether the moment-by-moment higher order
% correlations were reliably preserved across participants, we used
% inter-subject functional connectivity (ISFC) to isolate the
% time-varying correlational structure (functional connectivity patterns
% that were specifically driven by the story participants listened to.
% Following the analyses conducted by (HTFA)~\cite{MannEtal18}, we first
% applied \textit{hierarchical topographic factor analysis} (HTFA) to
% the fMRI datasets to obtain a time series of 700 node activities for
% every participant.  We then applied dimensionality reduction
% (Incremental PCA) for each group.

% We then compared the groups’ activity patterns (using Pearson
% correlations) to estimate the story times each corresponding pattern
% using more and more principle components.

% To assess decoding accuracy, we randomly divided participants for each
% stimulus into training and testing groups. We then compared the
% groups’ activity patterns (using Pearson correlations) to estimate the
% story times each corresponding pattern using more and more principle
% components (as the data became less compressed). Specifically, we asked, for each timepoint: what are the correlations
% between the first group's and second group's activity patterns at each
% order. We note that the decoding test we used is a conservative in which we count a timepoint label as incorrect if it is not an exact match.

\section*{Results}
By training classifiers using more and more principle components to decode, and comparing across conditions with varying degrees of cognitive richness, we can assess the explanatory power of the compressed data held with respect to the observed data (see \textit{Methods}). We note that our primary goal was not to achieve perfect decoding accuracy, but rather to use decoding accuracy as a benchmark for assessing whether different neural features specifically capture cognitively relevant brain patterns.

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{figs/decode_interpret.pdf}
  \caption{\textbf{Decoding accuracy.} \textbf{a. Decoding accuracy by
      number of components.} Ribbons of each color display
    cross-validated decoding performance for each condition (intact,
    paragraph, word, and rest). Decoders were trained using
    increasingly more principle components and displayed relative to
    chance (red line). \textbf{b. Fixed decoding accuracy by number of
      components.} We zoom in on the plot shown in \textbf{a.} and add
    a line denoting fixed decoding accuracy (.05). We plot where the
    intact, paragraph, and word conditions intersect.
    \textbf{c. Explanation of inflection metric.} First the we fit a sigmoid function to the decoding accuracy by number of components. Second, we found where the second derivative is both positive and less than .0001. Last, we then plot that inflection point as a single metric to capture the slope and asymptote of the curve.}
    \label{fig:decode_interpret}
  \end{figure}

Prior work has shown participants share similar neural responses to
richly structured stimuli when compared to stimuli with less
structure \cite{SimoEtal16}.  We replicate this finding, showing as complexity of the stimulus increases, decoding accuracy
increases (Fig.~\ref{fig:decode_interpret},  a.).  
Additionally, we found that as complexity of the stimuli increases, we need fewer components to decode the same amount (Fig.~\ref{fig:decode_interpret},  b.). However, we also found that as complexity of the stimuli increases, more components are required to reach peak decoding accuracy (Fig.~\ref{fig:decode_interpret},  c.).  
We posit that as the complexity of our thoughts increases, neural
compression decreases. However, as our thoughts become deeper and richer, more reliable information is available at higher neural compression.

% We performed a decoding analysis, using cross validation to estimate
% (using other participants’ data) which parts of the story the additional added
% principle component corresponded to (see \textit{Materials and methods}). We note that our primary goal was not to achieve perfect decoding accuracy, but rather to use decoding accuracy as a benchmark for assessing whether different neural features specifically capture cognitively relevant brain patterns.

% Separately for each experimental condition, we divided participants
% into two groups. Starting with 1 principle comonent, for
% each dimension we added another principle component, and we correlated the group 1 activity patterns with group 2
% activity patterns.  There were 272
% timepoints for paragraph condition, 300 timepoints for intact and word
% conditions, and 400 timepoints for rest condition,  so chance
% performance on this decoding test is was $\frac{1}{272}$,
% $\frac{1}{300}$, and $\frac{1}{400}$ respectively.

% - As complexity of the stimulus increases, decoding accuracy
% increases (Fig.~\ref{fig:decode_interpret},  a.).  Replication of
% Simony findings.

% - As complexity of the stimuli increases, we need fewer components to
% decode the same amount (Fig.~\ref{fig:decode_interpret},  b.).

% - As complexity of the stimuli increases, more components are required to reach peak decoding accuracy (Fig.~\ref{fig:decode_interpret},  c.).

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{figs/decode_pcs_network.pdf}
  \caption{\textbf{Inflection points by network.} \textbf{a.}
    Inflection point was calculated as explained in
    Fig.~\ref{fig:decode_interpret}, b. Analyses were limited by the
    brain networks (using the \cite{YeoEtal11} network parcellation) and
    arranged in increasing order relative to the intact condition.
    \textbf{b. and c.} For the total time in the intact condition, we are plotting the relative inflection points (\textbf{b.}) and corresponding number of components (\textbf{c.}) by network. \textbf{d.} The network parcellation defined by \cite{YeoEtal11} is displayed on the inflated brain maps. The colors and network labels serve as a legend for \textbf{a.} and \textbf{d.}}
  \label{fig:decode_pcs_network}
\end{figure}

We also wondered how this compression would change across brain regions.  We repeated the analysis but limited the brain hubs to 7 networks using the Yeo et al. (2011) network parcellation shown here in the inflated brain (Fig.~\ref{fig:decode_pcs_network}, d.). We found that as complexity of the stimuli increases, decoding accuracy increases
with higher cognitive areas. (Fig.~\ref{fig:decode_pcs_network}).

% As complexity of the stimuli increases, decoding accuracy increases
% with higher cognitive areas. (Fig.~\ref{fig: decode_pcs_network}).

%  If, there is some understanding of the narrative that accumulates over time, we should be able to see that change.

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{figs/decode_pcs.pdf}
  \caption{\textbf{Inflection points by thirds.} \textbf{a.}
    Decoding accuracy by number of components not broken into thirds
    (Fig.~\ref{fig:decode_interpret} a.). \textbf{b. and c.}
    Quantifying changes in decoding accuracy across
time. \textbf{b.} Slope of decoding accuracy was calculated by fitting
a regression line for each component/condition for each
third. \textbf{c.}  We also repeated the analysis
(Fig.~\ref{fig:decode_interpret}, b.) to obtain the inflection point
for each condition and for each third. \textbf{d.}  Decoding accuracy by number of components for each third of the scan time. We repeated the same analysis in Fig.~\ref{fig:decode_interpret} a. but breaking the scan time for each condition into 3 intervals.
}
  \label{fig:decode_pcs_thirds}
\end{figure}

We were also curious how compression would change across time.  If, there is some understanding of the narrative that accumulates over time, we should be able to see that difference. We found increases in decoding accuracy with the same number or fewer components for more complex, cognitively rich, conditions.
We also found decreases in decoding accuracy for the word-scrambled and rest
condition.

Overall, we found that as story listening conditions become more complex, more components are required to decode. We also found we could decode better with more impoverished data when
there is the underlying structure of the narrative providing more cognitive richness. We posit that as the complexity of our thoughts increases, neural compression decreases. However, as our thoughts become deeper and richer, more reliable information is available at higher neural compression.

% - Increases in decoding accuracy with the same number or fewer components for more complex, cognitively rich, conditions.
% - Decreases in decoding accuracy for the word-scrambled and rest
% condition.

@@

\section*{Author contributions}
Concept: J.R.M. and L.L.W.O. Implementation: L.L.W.O., and
J.R.M.  Analyses: L.L.W.O and J.R.M.

\bibliographystyle{apacite}
\bibliography{cdl.bib}
